Think deep for reef resilience

### 2025-09-25 paper/paper_draft.qmd
Shorten sentence
End on concrete management advice that summarises findings

Our results show increasing reef depth has a strong positive impact on modelled reef
resilience in temporal behaviour and maintenance of positive carbonate budgets under
possible future conditions. Reefs with a median depth of 10m or greater, with high incoming
connectivity, are the most likely to be assigned to a high cover time series cluster. As an
individual feature, incoming connectivity alone does not correlate with maintenance of
positive carbonate budgets. These findings highlight the importance of depth and incoming
connectivity as coupled selection features when assessing potential reef resilience under
future thermal stress.

@@

Our findings support the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and more prolonged resilience under climate change compared to
shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest that rubble
instability is inversely correlated with reef depth, possibly limiting juvenile coral
recruitment capacity of shallow reefs. A caveat to our depth-resilience findings is the
quality of the bathymetry data used to calculate median reef depth. As detailed previously,
estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution)
which can be subject to high degrees of uncertainty, both in terms of the estimated depth
for each pixel as well as their alignment to the established GBRMPA reef outlines.
Additionally, we do not incorporate light-based decreases in coral growth rates as depth
increases [@baker1975; @huston1985; @ruiz-diaz2022].

Current reef restoration and intervention efforts on the GBR, such as deployment of coral
seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs with
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, intervention depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1551
reefs), 59% (920) of reefs are consistently clustered into the Low and Medium cover clusters
across the GCMs (at the GBRMPA management region scale), with many unable to maintain
positive carbonate budgets under future conditions. Partial dependence analysis revealed a
range of depths from 10-15m where the probability of a reef being clustered into the High
cover cluster increased substantially. Additionally, two-way partial dependence analysis
revealed that high incoming connectivity is important for High cluster assignment for reefs
at depths of \>10m. Considering this 10-15m depth range when assessing possible intervention
sites may reduce the exposure of planted and seeded corals to high thermal stress, while
still maintaining effective logistical access to sites. Managers are also likely to consider
the potential for positive flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs around 15m deep. However, our results suggest no direct,
first order, relationship between weighted incoming connectivity and the ability of reefs to
maintain positive carbonate budgets under future conditions. There has been contrasting
evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling
is able to identify reefs that are important for recovery processes, with findings dependent
on the larval connectivity model used. These investigations were performed using prior
generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and
1996-2002 [@bode2018] time periods, without considering the influence of reef depth on
resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone may not be important or reef resilience and recovery, but should be considered in
combination with more influential factors such as depth. Comprehensive comparisons and
validation of multiple connectivity models, as well as use of a model ensemble, would
improve certainty and allow for the targetting of reefs by combining abiotic factors such as
depth with connectivity.

@@

>>>>

Our results show increasing reef depth has a strong positive impact on modelled reef
resilience in temporal behaviour and maintenance of positive carbonate budgets under
possible future conditions. Reefs with a median depth of 10m or greater, with high incoming
connectivity, are the most likely to be assigned to a high cover time series cluster. By
itself, incoming connectivity alone does not correlate with maintenance of positive
carbonate budgets. These findings highlight the importance of depth and incoming
connectivity as coupled selection features when assessing potential reef resilience under
future thermal stress.

@@

Our findings support the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and prolonged resilience (in terms of coral cover) under climate
change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024]
suggest that rubble instability is inversely correlated with reef depth, possibly limiting
juvenile coral recruitment capacity of shallow reefs. A caveat to our depth-resilience
findings is the quality of the bathymetry data used to calculate median reef depth. As
detailed previously, estimates of median reef depth are derived from satellite imagery (at
$10m^2$ resolution) which can be subject to high degrees of uncertainty, both in terms of
the estimated depth for each pixel as well as their alignment to the established GBRMPA reef
outlines. Additionally, we do not incorporate light-based decreases in coral growth rates as
depth increases [@baker1975; @huston1985; @ruiz-diaz2022].

Current reef restoration and intervention efforts across the GBR, such as deployment of
coral seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs
with sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there
are no explicit justifications for this depth range, intervention depth is likely
constrained by snorkelling and diving deployment and monitoring logistical accessibility,
visual identification of sites from the surface, in combination with tourism return value.
Our results suggest that of the assessed reefs occurring within the 1-10m depth range (1551
reefs), 59% (920) of reefs are consistently clustered into the Low and Medium cover clusters
across the GCMs (at the GBRMPA management region scale), with many found to be unable to
maintain positive carbonate budgets under projected conditions. Partial dependence analysis
revealed a range of depths from 10-15m where the probability of a reef being clustered into
the High cover cluster increased substantially. Additionally, two-way partial dependence
analysis revealed that high incoming connectivity is important for High cluster assignment
for reefs at depths of \>10m. Considering this 10-15m depth range when assessing possible
intervention sites may reduce the exposure of planted and seeded corals to high thermal
stress, while still maintaining effective logistical access to sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs around 15m deep. However, our results suggest no direct,
first order, relationship between weighted incoming connectivity and the ability of reefs to
maintain positive carbonate budgets under future conditions. There has been contrasting
evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling
is able to identify reefs that are important for recovery processes, with findings dependent
on the larval connectivity model used. These investigations were performed using prior
generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and
1996-2002 [@bode2018] time periods, without considering the influence of reef depth on
resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone may not be important or reef resilience and recovery, but should be considered in
combination with more influential factors such as depth. Comprehensive comparisons and
validation of multiple connectivity models, as well as use of a model ensemble, would
improve certainty and allow for the targeting of reefs by combining abiotic factors such as
depth with connectivity.

@@

Given the above, reef managers may consider the potential for positive flow-on effects due
to outgoing connectivity when selecting intervention sites. The necessary connectivity data,
however, must first be available to provide a sufficient degree of confidence to support
such site selection decisions (compare the ongoing discussion in [-@mumby2018] and
[-@bode2018]). Given connectivity is a transient, and much more difficult, property to
measure that is subject to comparatively higher sources of uncertainty, depth may be a more
reliable indicator of long-term resilience, particularly where no further information is
available.
